Museum specimens as baseline for assessing a potential global pollinator crisis

### 2018-03-26 ms.Rmd
More edits, specially to Spain analysis

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data is scarce and often geographically and taxonomically biased. This limits robust inference about any potential “pollinator crisis”. Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programms targetting pollinators are scarce. However, non-structured and often opportunistic historical data collections are more frequent. In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits). Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and even pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent, and that the lack of collection record data digitization is a key bottleneck, impeding progress in this field. We identify opportunities and limitations to progress in assessing pollinator declines at a global scale. Further, we demonstrate various approaches to analysing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. There is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally if we are able to make this data accesible and open.

@@

Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. This issue was identified several decades ago, and was popularized in 1996 by the influential book "the forgotten pollinators". However, these early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to asses the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010]. Yet the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011]. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017]. However, bees are also extremely diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification. The uneven distribution of bee researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos]. Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.

@@

Regardless of the method, it is important for researchers to go beyond aspects of biodiversity focused solely in species richness. In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015]. Environmental changes often results in few “winners” and many “losers” [@Bartomeus2013b]. Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018]. These changes can have important flow on effects for pollination of native plant species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

@@

[how much do we want to extend on this? Is interesting info, but boring o read?]()

@@

Start with butterflies?

@@

@@

Museum can provide valuable information besides species occurrence records, and this is important to take into account when digitizing the collections. This is specially relevant to identify mechanisms of decline and adaptation. One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change [#@Bartomeus2011]. In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected. This information is precious to understand past and present species interactions [#@Tylianakis2010ELE]. Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers]. Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.   
[move this section below]()

@@

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

@@

Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif. Upper right cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites. Lower right are sites which needs to dig museum records and lower left are countries hard to get data.

Alternative plots:  
```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)

@@

```
% makes it unidimensional. Simply logging the data may work, we can point out % of countries with > 1000 records in either axes, for example, etc... Correcting for country size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then?

Once museum collection datasets are made available, researchers needs to gauge any potential biases in the data. We explore this process with two contrasting examples from Spain and New Zealand. In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people. In this case, we contacted the original collectors to define their sampling protocols. We then resurveyed the same sites using the same protocols 20 years later. The New Zealand dataset includes a wide suite of collectors and collection locations. For this dataset, we ensured that there is no clear change to geographical [I haven’t actually looked at this as there weren’t and GPS coordinates in the data]() and taxonomic collection patterns through time. For example, some collectors may have targeted particular taxa, at specific locations or during certain time periods, which would represent a bias in the data. For these case studies, we provide annotated R scripts to analyse the different kinds of data (as described above). Essentially, these different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyze data for regions where there is little current information on pollinator declines. [include a repository where people can contribute?]

@@

In brief, we check sampling completeness at both time periods, and compare the richness level of each site before 1980 and after 1980 with a paired t-test. [Maybe more analyses arise...](). We show that ...

Fig 3: Multi-panel figure

@@

We thank E. Asensio (or maybe coauthor?), Curro Molina,  Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL. We also thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.

>>>>

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data is scarce and often geographically and taxonomically biased. This limits robust inference about any potential “pollinator crisis”. Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programms targetting pollinators are scarce. However, non-structured and often opportunistic historical data collections are more frequent. In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits). Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and even pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent, and that the lack of collection record data digitization is a key bottleneck, impeding progress in this field. We identify opportunities and limitations to progress in assessing pollinator declines at a global scale. Further, we demonstrate various approaches to analysing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. We show for both regions that despite richness levels do not change significantly, winners and looser species emerge. There is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally if we are able to make this data accesible and open.

@@

Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. This issue was identified several decades ago, and was popularized in 1996 by the influential book "the forgotten pollinators". However, these early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to asses the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010]. Yet, the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and includes different taxa sush as bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011]. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017]. However, bees are also extremely diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification. The uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos]. Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.

@@

In this paper we first highligth the main technical dificulties of using museum specimens, we show current evidence of diversity decline and highligth the need to go beyond diversity despriptors. Finally we draw a road map on how to exploit museum data to fill the knowledge gaps and ilustrate with two ase studies how to snalyze this kind of data.

@@

@@

Start with butterflies?
[Jamie, if you can complete this section with the refernces you found about declines would be super!]()

@@

#Plots look mainly white... I wonder if they are useful...

@@

[need integrate this two paragraphs!]()
Regardless of the method, it is important for researchers to go beyond aspects of biodiversity focused solely in species richness. A recent survey highlighted that more than 90% of the papers studying pollintar responses to land use change focused on richness and abundance descriptors. But in addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015]. Environmental changes often results in few “winners” and many “losers” [@Bartomeus2013b]. Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018]. These changes can have important flow on effects for pollination of native plant species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296

Museum can provide valuable information besides species occurrence records, and this is important to take into account when digitizing the collections. This is specially relevant to identify mechanisms of decline and adaptation. One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change [#@Bartomeus2011]. In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected. This information is precious to understand past and present species interactions [#@Tylianakis2010ELE]. Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers]. Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.

@@

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
#abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

@@

```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)

@@

```
Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif. Upper right cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites. Lower right are sites which needs to dig museum records and lower left are countries hard to get data.  
[we can point out % of countries with > 1000 records in either axes, for example, etc... would be cool to ggplot this and point where NZ and SP fall, for example. As well as the green and red dots in the lower right cuadrat]()

Once museum collection datasets are made available, researchers needs to gauge any potential biases in the data. We explore this process with two contrasting examples from Spain and New Zealand. In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people. In this case, we contacted the original collectors to define their sampling protocols. We then resurveyed the same sites using the same protocols 20 years later. The New Zealand dataset includes a wide suite of collectors and collection locations. For this dataset, we ensured that there is no clear change to geographical [I haven’t actually looked at this as there weren’t and GPS coordinates in the data]() and taxonomic collection patterns through time. For example, some collectors may have targeted particular taxa, at specific locations or during certain time periods, which would represent a bias in the data. For these case studies, we provide annotated R scripts to analyse the different kinds of data (as described above). Essentially, these different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyze data for regions where there is little current information on pollinator declines.

@@

In brief, we check sampling completeness at both time periods, and compare the richness level of each site before 1980 and after 1980 with a paired t-test. We show that overall there is no richness loss (p > 0.05), as 3 sites present lower number of species, but other three sites present higher numbers. However, raw numbers show a steep deline... and groups like nomadas, tyreus, prosopis and teralonias are ocmpletely extirpated (appendix?).

```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv("data/spain_clean.csv")
#head(dat)
library(reshape2)
dat2 <- melt(dat, id.vars = c("locality", "gen_sp", "when"), 
             measure.vars = "freq")
#head(dat2, 100)
#sort(unique(dat2$gen_sp))
#clean with clean_sp
#remove sp, NA, " ",
#NA_NA   
#Hylaeus_Variegatus

#table(dat2$gen_sp, dat2$when)
#Andrena_pilipes = carbonaria
#remove A. mellifera

#losts easely spotted: nomadas & Tyreus, prosopis, teralonias

#calculate richness per site and time period
#head(dat)
rich <- dcast(dat, locality ~ when, value.var = "gen_sp", 
              fun.aggregate = function(x) length(unique(x)))
#rich
#t.test(rich$asensio, rich$now, paired = TRUE)

library(vegan)
temp <- subset(dat, when == "now")
comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
              fun.aggregate = sum, na.rm = TRUE)
rownames(comm)[-8] <- as.character(comm$locality[-8])
comm2 <- comm[,-1]
rowSums(comm2)
temp <- specaccum(comm2, "random")
#plot(temp)
#rarecurve(comm2)
now <- rarefy(x = comm2, sample = 120)

temp <- subset(dat, when == "asensio")
comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
              fun.aggregate = sum, na.rm = TRUE)
rownames(comm)[-8] <- as.character(comm$locality[-8])
comm2 <- comm[,-1]
rowSums(comm2)
temp <- specaccum(comm2, "random")
#plot(temp)
#rarecurve(comm2)
asensio <- rarefy(x = comm2, sample = 120)

ttest <- t.test(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)], paired = TRUE)

boxplot(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)], names = c("1980's", "2016"), ylab = "rarefies species richness", 
            las = 1, border = "light grey")
points(x = c(rep(1,5), rep (2,5)), y = c(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)]), col = rep(1:5,2), pch = 19)
for(i in c(1,2,5,6,7)){
    lines(x = 1:2, y = c(asensio[i], now[i]), col = i)
}

```
Fig 3: Rarefied richness at the 5 plots... need to work more on the raw data.

@@

IB was funded by FBBVA project. We thank E. Asensio for sharing his histrical data and knowledge. We thank Curro Molina, Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum of Historia Natural and ITACyL for access to historical collections. We also thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.
